---
title: "Colony Mice"
author: "Michael McCoy"
logo: "https://yt3.googleusercontent.com/ytc/APkrFKbprBkcDrFaaYruHcnwzSgxUJV2N0799D788ar6dQ=s900-c-k-c0x00ffffff-no-rj"
nav-buttons:
- icon: github
  href: https://github.research.chop.edu/mmccoy-01?tab=repositories
- icon: linkedin
  href: https://www.linkedin.com/in/themichaelmccoy
format: 
  dashboard:
    orientation: columns
    theme:
      light: cerulean
      dark: darkly
---

```{r Setup}
#| context: setup

# Set the working directory to the location of the current script. Use this only when running code chunks interactively, not when rendering the entire document.
#setwd(dirname(rstudioapi::getSourceEditorContext()$path))

# Do not run this code or update the renv.lock file, as it reflects the original state when the code was written and executed.
#renv::snapshot(lockfile = "renv.lock")

# Do not run this code or update the manifest.json file, as it reflects the original state when the code was written and executed.
#rsconnect::writeManifest()

# List of packages
pkg_list <- c(
  "knitr", "tidyverse", "shiny", "magick", "rsvg", "DT",
  "scales", "plotly", "survival", "survminer", "jmv", "igraph", "ggraph", "cli"
)

# Function to install and load packages
load_packages <- function(pkg_list) {
  for (pkg in pkg_list) {
    if (!require(pkg, character.only = TRUE)) {
      install.packages(pkg, dependencies = TRUE)
      library(pkg, character.only = TRUE)
    }
  }
}

load_packages(pkg_list)

# List of current experiment file paths (pointing directly to specific CSV files)
file_paths <- c(
)

# Function to read and combine specific columns from data frames
read_and_combine_selected_data <- function(paths) {
  # Initialize an empty data frame with the desired columns
  combined_data <- data.frame(id = character(), 
                               experiment = character(), 
                               cage_number = character(), 
                               trt = character(), 
                               dob = character(),
                               sex = character(),
                               manner_of_death = character(),  
                               stringsAsFactors = FALSE)
  
  # Loop through each file path
  for (path in paths) {
    new_data <- read.csv(path)  # Read the CSV file
    
    # Ensure that all relevant columns are present
    new_data <- new_data %>%
      mutate(across(c(id, cage_number, dob, manner_of_death), 
                    ~ as.character(.), 
                    .names = "{.col}"))  # Convert specified columns to character type

    # Check if 'trt' exists and convert if so, otherwise fill with NA
    if ("trt" %in% colnames(new_data)) {
      new_data <- new_data %>%
        mutate(trt = as.character(trt))
    } else {
      new_data$trt <- NA  # Fill 'trt' with NA if it doesn't exist
    }
    
    # Select only the columns of interest
    new_data_combined <- new_data %>%
      select(id, experiment, cage_number, trt, dob, sex, manner_of_death)
    
    # Bind the new data frame to the combined data
    combined_data <- bind_rows(combined_data, new_data_combined)
  }
  
  return(combined_data)
}

# Combine all data using the specified columns
raw_experiment_data <- read_and_combine_selected_data(file_paths)

# Filter for unique rows by id and experiment so as to have wide data and not long data
raw_experiment_data <- raw_experiment_data %>% 
  distinct(id, experiment, .keep_all = TRUE)

# Calculate number of breeder cages
breeder_cage_count <- read.csv("data/raw/raw_data.csv") %>%
  filter(is_active != "n",
         status == "breeder") %>% 
  summarize(unique_cages = n_distinct(cage_number))

# Calculate number of NSG-Hc1 colony cages
nsghc1_colony_cage_count <- read.csv("data/raw/raw_data.csv") %>%
  filter(is_active != "n",
         status == "weanling",
         strain == "NSG-Hc1") %>% 
  summarize(unique_cages = n_distinct(cage_number))

# Calculate number of NSG colony cages
nsg_colony_cage_count <- read.csv("data/raw/raw_data.csv") %>%
  filter(is_active != "n",
         status == "weanling",
         strain == "NSG") %>% 
  summarize(unique_cages = n_distinct(cage_number))

# Calculate number of experiment cages
experiment_cage_count <- raw_experiment_data %>% 
  filter(is.na(manner_of_death) | manner_of_death == "") %>% 
  summarize(unique_cages = n_distinct(cage_number))

# Count distinct combinations of id and experiment where manner_of_death is NA
experiment_mice <- raw_experiment_data %>%
  filter(is.na(manner_of_death) | manner_of_death == "") %>%
  distinct(id, experiment) %>%
  summarise(n_unique_mice = n())

# Calculate number of expansion cages
expansion_cage_count <- read.csv("data/raw/raw_data.csv") %>%
  filter(is_active != "n",
         status == "expansion") %>% 
  summarize(unique_cages = n_distinct(cage_number))
```

# Colony

## Column

### Row

#### Breeders {.tabset}

##### Breeders

```{r Breeders}
#|title: "Breeders"

# An optimal age for pairing a sire and a dam is:
# Sire: 8 to 12 weeks old; Retire at 52 weeks
# Dam: 8 to 10 weeks old; Retire at 6-8 months or about 26 to 35 weeks

# Load data
raw_data <- read.csv("data/raw/raw_data.csv")

# Select the desired columns
raw_data <- raw_data %>%
  filter(is_active != "n",
         status == "breeder",
         weaned == "") %>% 
  select(-status,
         -is_active,
         -sex,
         -room,
         -strain,
         -num_of_mice_in_cage,
         -weaned,
         -parent_cage,
         -notes,
         -expansion,
         -injection_date,
         -num_of_cells_per_mouse) %>% 
  arrange(cage_number)

raw_data <- raw_data %>%
  mutate(
    # Split dob into sire and dam dob
    sire_dob = sapply(strsplit(as.character(raw_data$dob), ","), `[`, 1),
    dam_dob = sapply(strsplit(as.character(raw_data$dob), ","), `[`, 2),
    
    # Convert to Date format using ymd (for "YYYY-MM-DD" format)
    sire_dob = ymd(sire_dob),
    dam_dob = ymd(dam_dob),
    
    # Calculate weeks old or set to 'unknown' only if both are missing
    weeks_old = case_when(
      is.na(sire_dob) & is.na(dam_dob) ~ "unknown",
      is.na(sire_dob) ~ paste("unknown,", floor(difftime(Sys.Date(), dam_dob, units = "weeks"))),
      is.na(dam_dob) ~ paste(floor(difftime(Sys.Date(), sire_dob, units = "weeks")), ", unknown"),
      TRUE ~ paste(
        floor(difftime(Sys.Date(), sire_dob, units = "weeks")),
        floor(difftime(Sys.Date(), dam_dob, units = "weeks")),
        sep = ", "
      )
    )
  ) %>% 
  select(weeks_old, everything(),
         -dob,
         -sire_dob,
         -dam_dob)

# Create an interactive DataTable
datatable(raw_data)
```

##### Parent-Weanling Relationship

```{r Parent-Weanling Relationship}
#|title: "Parent-Weanling Relationship"

# Load the data
raw_data <- read.csv("data/raw/raw_data.csv")

# Filter for weanlings and breeders and add weeks_old
raw_data <- raw_data %>%
  filter((status == "weanling" | status == "breeder") & 
           is_active == "" & 
           strain == "NSG-Hc1") %>% 
  mutate(weeks_old = round(as.numeric(difftime(Sys.Date(), as.Date(dob), units = "weeks")), 1))

# Filter out rows where parent_cage is NA (i.e., root of the family tree)
edges <- raw_data %>%
  filter(!is.na(parent_cage)) %>%
  select(parent_cage, cage_number, sex)

# Handle mixed parent_cage values (split by comma)
edges <- edges %>%
  separate_rows(parent_cage, sep = "\\s*,\\s*")  # Split on comma and trim spaces

# Create the graph object
graph <- graph_from_data_frame(edges, directed = TRUE)

# Create a data frame with cage_number for each node
node_cage_numbers <- unique(c(edges$parent_cage, edges$cage_number))
node_data <- data.frame(name = node_cage_numbers, cage_number = node_cage_numbers)

# Add sex information to node data
node_data$sex <- ifelse(node_data$name %in% edges$cage_number[edges$sex == "male"], "male", "female")

# Add cage_number and sex as node attributes
V(graph)$cage_number <- node_data$cage_number[match(V(graph)$name, node_data$name)]
V(graph)$sex <- node_data$sex[match(V(graph)$name, node_data$name)]

# Get node coordinates for a tree layout
layout <- layout_as_tree(graph)

# Create a data frame for nodes with layout coordinates and sex
node_positions <- data.frame(
  x = layout[, 1],
  y = layout[, 2],
  name = V(graph)$name,
  cage_number = V(graph)$cage_number,
  sex = V(graph)$sex
)

# Create a data frame for edges with correct coordinates
edge_positions <- data.frame(
  x = node_positions$x[match(edges$parent_cage, node_positions$name)],
  y = node_positions$y[match(edges$parent_cage, node_positions$name)],
  xend = node_positions$x[match(edges$cage_number, node_positions$name)],
  yend = node_positions$y[match(edges$cage_number, node_positions$name)]
)

# Define colors based on sex
node_positions$color <- ifelse(node_positions$sex == "male", "lightblue", "lightpink")

# Convert both cage_number columns to character and trim spaces
node_positions$cage_number <- as.character(trimws(node_positions$cage_number))
raw_data$cage_number <- as.character(trimws(raw_data$cage_number))

# Perform the left join to include num_of_mice_in_cage and weeks_old
node_positions <- node_positions %>%
  left_join(
    raw_data %>%
      select(cage_number, num_of_mice_in_cage, weeks_old), 
    by = "cage_number"
  )

# Create the Plotly graph
plot <- plot_ly() %>%
  # Add edges as lines
  add_segments(
    data = edge_positions,
    x = ~x, y = ~y,
    xend = ~xend, yend = ~yend,
    line = list(color = 'black', width = 1),
    hoverinfo = 'none'
  ) %>%
  # Add nodes as markers
  add_trace(
    data = node_positions,
    x = ~x, y = ~y,
    type = 'scatter',
    mode = 'markers+text',
    text = ~cage_number,
    textposition = 'top center',
    marker = list(size = 10, color = ~color),
    hoverinfo = "text",
    hovertext = ~paste(
      "Cage: ", cage_number, "<br>",
      num_of_mice_in_cage, " mice in cage<br>",
      "Weeks old: ", weeks_old
    )
  ) %>%
  layout(
    title = 'Mouse Colony Family Tree',
    xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
    yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE)
  )

# Display the plot
plot
```

### Column

#### Column

There are `{r} breeder_cage_count` breeder cages as of `r format(Sys.Date(), "%A, %B %d, %Y")`.

```{r Litters}
#| title: "Litters"

# Load data
raw_data <- read.csv("data/raw/raw_data.csv") %>%
  filter(is_active != "n",
         status == "breeder",
         dob_litter != "",
         weaned == ""
  )

# Calculate the difference in days between today's date and the `wean_date`
today <- Sys.Date()

# Function to generate a progress bar based on days_to_wean
generate_progress_bar <- function(days_to_wean, total_days = 21) {
  # Handle cases where days_to_wean is negative
  if (days_to_wean < 0) {
    return("[====================] (Wean date passed)")
  }
  
  # Cap the progress bar range
  completed_ratio <- max(0, min(1, (total_days - days_to_wean) / total_days))
  progress_length <- 20
  filled_length <- round(progress_length * completed_ratio)
  empty_length <- progress_length - filled_length
  
  filled <- strrep("=", filled_length)
  empty <- strrep("-", empty_length)
  
  paste0("[", filled, empty, "]")
}

# Convert `wean_date` to Date format and apply calculation
raw_data <- raw_data %>%
  arrange(cage_number) %>% 
  mutate(wean_date = as.Date(wean_date, format = "%Y-%m-%d"),
         days_to_wean = as.numeric(wean_date - today),
         message = ifelse(days_to_wean < 0,
                          paste0("Cage ", cage_number, 
                                 " has ", litter_size, 
                                 " pups that should have been weaned by ", wean_date, "."),
                          paste0("Cage ", cage_number, 
                                 " has ", litter_size, 
                                 " pups ready to be weaned in ", days_to_wean, 
                                 " days on ", wean_date, ".")),
         progress_bar = map_chr(days_to_wean, generate_progress_bar))

# Print messages with progress bars using pmap and suppress output
invisible(pmap(raw_data, function(message, progress_bar, ...) {
  cat(message, "\n")
  cat(progress_bar, "\n\n")
}))
```

## Column

### Weanlings

```{r Weanlings}
#|title: "Weanlings"

# Load data
raw_data <- read.csv("data/raw/raw_data.csv")

# Select the desired columns and add a weeks_old column
raw_data <- raw_data %>%
  filter(is_active != "n",
         status == "weanling"
         ) %>%
  select(-status,
         -is_active,
         -introduced_date,
         -dob_litter,
         -litter_size,
         -wean_date,
         -num_of_litters,
         -weaned,
         -parent_cage,
         -injection_date,
         -num_of_cells_per_mouse,
         -expansion
) %>% 
  mutate(weeks_old = round(as.numeric(difftime(Sys.Date(), as.Date(dob), units = "weeks")), 1)) %>% 
  relocate(weeks_old, .after = dob) %>% 
  arrange(strain, cage_number)

# Create an interactive DataTable
datatable(raw_data, 
          extensions = 'Buttons',
          options = list(
            pageLength = 20,
            scrollX = TRUE,
            dom = 'Bfrtip',  # Buttons at the top
            buttons = c('csv')  # CSV download button
          )
)
```

```{r Colony Mice}
#| title: "Colony mice"

current_date <- format(Sys.Date(), "%A, %B %d, %Y")

# Create first bullet point for NSG-Hc1 and NSG colony cages
colony_message <- paste0(
  "There are ", nsghc1_colony_cage_count, 
  " NSG-Hc1 colony cages and ", nsg_colony_cage_count, 
  " NSG colony cages as of ", current_date, "."
)

# NSG-Hc1 Mice Code:
nsg_hc1_message <- read.csv("data/raw/raw_data.csv") %>%
  filter(is_active != "n",
         status == "weanling",
         strain == "NSG-Hc1") %>%
  mutate(dob = as.Date(dob, format = "%Y-%m-%d"),
         days_old = as.numeric(Sys.Date() - dob)) %>%
  group_by(status, sex) %>%
  summarise(
    total_mice_by_sex = sum(num_of_mice_in_cage, na.rm = TRUE), 
    avg_days_old = mean(days_old, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  summarize(
    total_mice = sum(total_mice_by_sex, na.rm = TRUE),
    female = sum(total_mice_by_sex[sex == "female"], na.rm = TRUE),
    male = sum(total_mice_by_sex[sex == "male"], na.rm = TRUE),
    weighted_avg_days_old = sum(total_mice_by_sex * avg_days_old) / total_mice,
    avg_weeks_old = round(weighted_avg_days_old / 7, 1)
  ) %>%
  mutate(
    message = paste0("You have ", total_mice, 
                     " NSG-Hc1 mice (", female, " female and ", male, 
                     " male) that are about ", avg_weeks_old, " weeks old.")
  ) %>%
  pull(message)

# NSG Mice Code:
nsg_message <- read.csv("data/raw/raw_data.csv") %>%
  filter(is_active != "n",
         status == "weanling",
         strain == "NSG") %>%
  mutate(dob = as.Date(dob, format = "%Y-%m-%d"),
         days_old = as.numeric(Sys.Date() - dob)) %>%
  group_by(status, sex) %>%
  summarise(
    total_mice_by_sex = sum(num_of_mice_in_cage, na.rm = TRUE), 
    avg_days_old = mean(days_old, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  summarize(
    total_mice = sum(total_mice_by_sex, na.rm = TRUE),
    female = sum(total_mice_by_sex[sex == "female"], na.rm = TRUE),
    male = sum(total_mice_by_sex[sex == "male"], na.rm = TRUE),
    weighted_avg_days_old = sum(total_mice_by_sex * avg_days_old) / total_mice,
    avg_weeks_old = round(weighted_avg_days_old / 7, 1)
  ) %>%
  mutate(
    message = paste0("You have ", total_mice, 
                     " NSG mice (", female, " female and ", male, 
                     " male) that are about ", avg_weeks_old, " weeks old.")
  ) %>%
  pull(message)
```

- `r colony_message`
- `r nsg_hc1_message`
- `r nsg_message`

# Studies

## Row

There are `r experiment_cage_count` experiment cages and `r experiment_mice` mice as of `r format(Sys.Date(), "%A, %B %d, %Y")`.

### Column

```{r Study}
#|title: "Study Mice"

# Select the desired columns and add a weeks_old column
raw_data <- raw_experiment_data %>%
  mutate(weeks_old = round(as.numeric(difftime(Sys.Date(), as.Date(dob), units = "weeks")), 1)) %>% 
  relocate(weeks_old, .after = dob)

# Count distinct combinations of id and experiment where manner_of_death is NA
experiment_mice <- raw_experiment_data %>%
  distinct(id, experiment) %>%
  summarise(n_unique_mice = n())

# Create an interactive DataTable
datatable(raw_data)
```

```{r Current Status by Cage and Treatment}
#| title: "Current Status"
#| results: asis

# Extract unique experiments
unique_experiments <- unique(raw_experiment_data$experiment)

# Loop through each unique experiment and create plots
for (exp in unique_experiments) {
  scatterplot <- raw_experiment_data %>%
    # Check if 'trt' column exists, if not create it with default value "Unknown"
    {if (!"trt" %in% names(.)) mutate(., trt = "Unknown") else .} %>%
    arrange(manner_of_death) %>%  # Arrange data to ensure NA is at the end
    distinct(trt, id, experiment, .keep_all = TRUE) %>%  # Ensure uniqueness by experiment
    # Replace missing trt values with "Unknown"
    mutate(trt = ifelse(is.na(trt) | trt == "", "Unknown", trt)) %>%
    # Ensure id is numeric (if it should be), otherwise, use factor levels to reorder
    mutate(id_numeric = as.numeric(id)) %>%  # Create a numeric version of id
    filter(experiment == exp) %>%  # Filter data for the current experiment
    ggplot(aes(x = factor(cage_number), y = reorder(trt, id_numeric, na.rm = TRUE))) +  # Reorder using numeric id
    geom_point(
      aes(color = factor(ifelse(is.na(manner_of_death) | manner_of_death == "", "Alive", "Dead"))),
      size = 3,
      position = position_dodge2(width = 0.5),
      alpha = 0.7
    ) +
    geom_text(
      aes(label = as.character(id), color = factor(ifelse(is.na(manner_of_death) | manner_of_death == "", "Alive", "Dead"))),
      position = position_dodge2(width = 0.5),
      vjust = -1,
      size = 3
    ) +
    scale_color_manual(
      values = c("Dead" = "red", "Alive" = "green"),
      guide = guide_legend(title = "Status")
    ) +
    labs(x = "Cage Number", y = "Treatment", color = "Status") +
    theme_minimal() +
    theme(legend.position = "top", 
          axis.text.x = element_text(size = 12, margin = margin(0, 40, 0, 40))) +
    guides(color = "none") +
    scale_x_discrete(breaks = unique(raw_experiment_data$cage_number)) +
    ggtitle(paste("Scatterplot for Experiment:", exp))  # Title for the plot

  # Print the scatterplot for the current experiment
  print(scatterplot)
}
```

# Expansions

## Column

There are `{r} expansion_cage_count` expansion cages as of `r format(Sys.Date(), "%A, %B %d, %Y")`.

```{r Expansions}
#|title: "Expansions"

# Load data
raw_data <- read.csv("data/raw/raw_data.csv")

# Select the desired columns
raw_data <- raw_data %>%
  filter(is_active != "n",
         status == "expansion") %>% 
  select(-status,
         -is_active,
         -introduced_date,
         -dob_litter,
         -litter_size,
         -num_of_litters,
         -wean_date,
         -weaned,
         -parent_cage) %>% 
  mutate(weeks_old = round(as.numeric(difftime(Sys.Date(), as.Date(dob), units = "weeks")), 1),
         days_since_injection = round(as.numeric(difftime(Sys.Date(), as.Date(injection_date), units = "days")), 1),
         num_of_cells_per_mouse = formatC(num_of_cells_per_mouse, format = "e", digits = 2)
         ) %>% 
  relocate(weeks_old, .after = dob) %>% 
  arrange(cage_number)

# Create an interactive DataTable
datatable(raw_data, 
          extensions = 'Buttons',
          options = list(
            pageLength = 20,
            scrollX = TRUE,
            dom = 'Bfrtip',  # Buttons at the top
            buttons = c('csv')  # CSV download button
          )
)
```


# CAGE CARD GENERATION TEST CODE

```{r CAGE CARD GENERATION TEST CODE}
#| eval: false

# Load necessary libraries
library(ggplot2)
library(gridExtra)
library(qrcode)
library(baRcodeR)
library(grid)
library(magick)

# Example data
cage_data <- data.frame(
  mouse_id = 1:5,
  cage_number = rep("Cage 101", 5),
  mouse_line = rep("Line A", 5),
  strain = rep("NSG | 005557", 5),
  sex = c("M", "F", "M", "M", "F"),
  generation = rep("F1", 5),
  PI = rep("Grupp", 5),
  technician = rep("Michael", 5),
  phone = rep("(610) 909-8680", 5),
  born = c("2024-01-10", "2024-01-15", "2024-01-20", "2024-01-25", "2024-02-01"),
  genotype = rep("WT", 5),
  gene = rep("GeneX", 5),
  additional_notes = rep("No issues", 5)
)

# Remove duplicate cage_number rows (ensure only one card per cage_number)
cage_data_unique <- cage_data[!duplicated(cage_data$cage_number), ]

# Function to create each cage card without QR code
generate_cage_card <- function(data_row) {
  ggplot() +
    geom_text(aes(x = 0, y = 5.5), label = paste("Cage:", data_row$cage_number), hjust = 0, size = 5) +
    geom_text(aes(x = 0, y = 5), label = paste("Mouse Line:", data_row$mouse_line), hjust = 0, size = 4) +
    geom_text(aes(x = 0, y = 4.5), label = paste("Strain:", data_row$strain), hjust = 0, size = 4) +
    geom_text(aes(x = 0, y = 4), label = paste("Sex:", data_row$sex), hjust = 0, size = 4) +
    geom_text(aes(x = 0, y = 3.5), label = paste("Generation:", data_row$generation), hjust = 0, size = 4) +
    geom_text(aes(x = 0, y = 3), label = paste("PI:", data_row$PI), hjust = 0, size = 4) +
    geom_text(aes(x = 0, y = 2.5), label = paste("Technician:", data_row$technician), hjust = 0, size = 4) +
    geom_text(aes(x = 0, y = 2), label = paste("Phone:", data_row$phone), hjust = 0, size = 4) +
    
    # Blue Circle Sticker
    annotate("point", x = 11, y = 5, color = "blue", size = 8) +
    
    # Aesthetic adjustments
    theme_void() +
    theme(plot.margin = margin(1, 1, 1, 1, "cm"))
}

# Generate cage card for each unique cage number
cage_cards <- lapply(seq_len(nrow(cage_data_unique)), function(i) {
  generate_cage_card(cage_data_unique[i, ])
})

# Set the dimensions of the cage card (12.6 cm L by 6 cm H)
pdf("Cage_Cards.pdf", width = 12.6 / 2.54, height = 6 / 2.54)  # Convert cm to inches
do.call(grid.arrange, c(cage_cards, ncol = 1))  # Arrange one card per page
dev.off()
```